
# AbsPQN

Loading packages and NPX data.

```{r}
# Packages + AbsPQN function
library(tidyverse)
library(OlinkAnalyze)
library(ggplot2)
library(patchwork)
library(umap)
source("norm_AbsPQN.R")

# Data, undergone QC so samples and proteins have been removed
npx_skane_raw <- read_csv("../clean_data/npx_skane_filtered.csv", show_col_types = FALSE) 
npx_sthlm_raw <- read_csv("../clean_data/npx_sthlm_filtered.csv", show_col_types = FALSE)

binfo <- read.csv("protein_information.csv")

clinical_skane <- read_csv("../clean_data/clinical_skane_filtered.csv", show_col_types = FALSE)
clinical_sthlm <- read_csv("../clean_data/clinical_sthlm_filtered.csv", show_col_types = FALSE)
```

To avoid normalizing the sample_ids, they are kept as rownames instead of a separate column.

```{r}
# Converting sample_ids to rownames
npx_skane <- npx_skane_raw |> 
  arrange(sample_id) |> 
  column_to_rownames(var = "sample_id")  

npx_sthlm <- npx_sthlm_raw |>  
  arrange(sample_id) |> 
  column_to_rownames(var = "sample_id")
```

We will use two types of AbsPQN: The first is done by simply inputting the data in its current format (log2) and using the output directly, let's call this AbsPQN1 (v1). Secondly, AbsPQN2 (v2) is done by removing log2 by taking the inverse log transformation (2^), applying AbsPQN, then reapplying log2 to the data.

```{r}
# Kept in log2(data) format
npx_skane_v1 <- npx_skane
npx_sthlm_v1 <- npx_sthlm

# Removing log2
npx_skane_v2 <- 2^(npx_skane)
npx_sthlm_v2 <- 2^(npx_sthlm)
```

Now, we apply the AbsPQN function on the data sets. *Note!* AbsPQN cannot take tibbles as input, therefore, they are converted into matrices.

```{r}
# AbsPQN1
result_skane_v1 <- abspqn(as.matrix(npx_skane_v1), by_s = NULL, by_b = NULL) 
result_sthlm_v1 <- abspqn(as.matrix(npx_sthlm_v1), by_s = NULL, by_b = NULL)

# AbsPQN2
result_skane_v2 <- abspqn(as.matrix(npx_skane_v2), by_s = NULL, by_b = NULL) |> 
  log2()
result_sthlm_v2 <- abspqn(as.matrix(npx_sthlm_v2), by_s = NULL, by_b = NULL) |> 
  log2()
```

Lastly, before outputting the data, the data frames need to be cleaned up.

```{r}
# AbsPQN1
norm_npx_skane_v1 <- result_skane_v1 |> 
  as_tibble(rownames = NA) |>  # Converted back into a tibble
  mutate(sample_id = rownames(result_skane_v1)) |> # Restores the sample_ids as a column, from the row names
  select(sample_id, everything()) # Reorders the columns so that sample_id is first

norm_npx_sthlm_v1 <- result_sthlm_v1 |> 
  as_tibble(rownames = NA) |> 
  mutate(sample_id = rownames(result_sthlm_v1)) |> 
  select(sample_id, everything())

# AbsPQN2
norm_npx_skane_v2 <- result_skane_v2 |> 
  as_tibble(rownames = NA) |> 
  mutate(sample_id = rownames(result_skane_v2)) |> 
  select(sample_id, everything())

norm_npx_sthlm_v2 <- result_sthlm_v2 |> 
  as_tibble(rownames = NA) |> 
  mutate(sample_id = rownames(result_sthlm_v2)) |> 
  select(sample_id, everything())
```

Saving the data sets as csv files.

```{r}
# Kept in log2
write_csv(norm_npx_skane_v1, file.path("../clean_data/norm_npx_skane_v1.csv"))
write_csv(norm_npx_sthlm_v1, file.path("../clean_data/norm_npx_sthlm_v1.csv"))

# Removed log2
write_csv(norm_npx_skane_v2, file.path("../clean_data/norm_npx_skane_v2.csv"))
write_csv(norm_npx_sthlm_v2, file.path("../clean_data/norm_npx_sthlm_v2.csv"))
```

# Visualization

## Olink IQR vs. Median plot

To evaluate the effect of AbsPQN, let's reexamine the IQR vs. Median plot, supplied by the OlinkAnalyze package. Before creating the plots, some extra data is needed and the NPX data sets need to be converted into long format (for plotting).

```{r}
# Retrieving necessary info about panels and proteins
olink_ids <- data.frame(Protein = binfo$column_name, OlinkID = binfo$olink_id) # OlinkID
panel_ids <- data.frame(Protein = binfo$column_name, Panel = binfo$panel)      # Type of panel  

# Creating function for adding info, converting to long, and plotting IQR vs. Median plot
olink_plot_maker <- function(df, plot_title) {
  
  # Converting to long, then adding information 
  long_df <- pivot_longer(df, cols = -sample_id, names_to = "Protein", values_to = "NPX") |> 
    left_join(olink_ids, by = "Protein") |> 
    left_join(panel_ids, by = "Protein") |> 
    rename(SampleID = sample_id) |> 
    mutate(SampleID = as.character(SampleID))
  
  # Plot
  IQR_median_plot<- long_df |> 
  olink_qc_plot(color_g = "Panel", IQR_outlierDef = 5, median_outlierDef = 5) +
    labs(title = plot_title)
  
  return(IQR_median_plot)
}
  
# Calling the function
IQR_a <- olink_plot_maker(norm_npx_skane_v1, "Skane - AbsPQN1")
IQR_b <- olink_plot_maker(norm_npx_sthlm_v1, "Stockholm - AbsPQN1")
IQR_c <- olink_plot_maker(norm_npx_skane_v2, "Skane - AbsPQN2")
IQR_d <- olink_plot_maker(norm_npx_sthlm_v2, "Stockholm - AbsPQN2")
```

Plotting..

```{r}
# Skane
IQR_a
IQR_c

# Sthlm
IQR_b
IQR_d 
```

## UMAP

Here, a function for running UMAP, extracting the results, and plotting them is created.

```{r}
# Creating function for both running, extracting, and plotting UMAP results
umap_plot <- function(df, plot_title) {
  
  # Running UMAP
  umap_df <- df |> 
    select(-sample_id) |> 
    umap()
  
  # Extracting UMAP variables
  df$UMAP1 <- umap_df$layout[, 1]
  df$UMAP2 <- umap_df$layout[, 2]
  
  # Making UMAP plot
  plot <- ggplot(df, aes(x = UMAP1, y = UMAP2, color = protein_name))
    geom_point(alpha = 0.8) +
    labs(title = plot_title, x = "UMAP 1", y = "UMAP 2") +
    theme_minimal() +
    scale_color_viridis_c()
  
  return(plot)
}
```

Now, we run the function to run UMAP and generate the plots for each of the data sets.

```{r}
# Using the function to create a plot for each data set
umap_a <- umap_plot(norm_npx_skane_v1, "Skane - AbsPQN1")
umap_b <- umap_plot(norm_npx_sthlm_v1, "Stockholm - AbsPQN1")
umap_c <- umap_plot(norm_npx_skane_v2, "Skane - AbsPQN2")
umap_d <- umap_plot(norm_npx_sthlm_v2, "Stockholm - AbsPQN2")
```

Plotting..

```{r}
umap_a + umap_c
umap_b + umap_d
```