```{r}
library(tidyverse)
library(viridis)
library(FactoMineR)
library(factoextra)
library(pheatmap)
library(umap)
library(patchwork)
library(limma)
library(car)
library(ggridges)
library(ggsci)

source("association_functions.R")
```

```{r}
# Loading NPX- and clinical data
npx_skane <- read_csv("../clean_data/norm_npx_skane_v2.csv", show_col_types = FALSE) # Using NPX which removed log2 during AbsPQN
clinical_skane <- read_csv("../clean_data/clinical_skane_filtered.csv", show_col_types = FALSE)

npx_sthlm <- read_csv("../clean_data/norm_npx_sthlm_v2.csv", show_col_types = FALSE)
clinical_sthlm <- read_csv("../clean_data/clinical_sthlm_filtered.csv", show_col_types = FALSE)

npx_raw <- read.csv("../data/2024-01-23_karma_pro_npx.csv") # No QC
sinfo_raw <- read.csv("../data/2024-01-23_karma_pro_sinfo.csv") # No QC
```

```{r}
# Can only use clinical variables without missing values, thus, here the NPX data will be filtered to only include samples with complete variables

# Alcohol
clinical_skane_alcohol <- clinical_skane |> 
  filter(!is.na(alcohol_gram_week))
npx_skane_alcohol <- npx_skane |> 
  filter(sample_id %in% clinical_skane_alcohol$sample_id)

clinical_sthlm_alcohol <- clinical_sthlm |> 
  filter(!is.na(alcohol_gram_week))
npx_sthlm_alcohol <- npx_sthlm |> 
  filter(sample_id %in% clinical_sthlm_alcohol$sample_id)

# Smoking
clinical_skane_smoking<- clinical_skane |> 
  filter(!is.na(smoking_status))
npx_skane_smoking <- npx_skane |> 
  filter(sample_id %in% clinical_skane_smoking$sample_id)

clinical_sthlm_smoking<- clinical_sthlm |> 
  filter(!is.na(smoking_status))
npx_sthlm_smoking <- npx_sthlm |> 
  filter(sample_id %in% clinical_sthlm_smoking$sample_id)

# Hormonal treatment status: 2 = current, 1 = previous, 0 = never
clinical_skane_hrt <- clinical_skane |> 
  filter(!is.na(mht_status))
npx_skane_hrt <- npx_skane |> 
  filter(sample_id %in% clinical_skane_hrt$sample_id)

clinical_sthlm_hrt <- clinical_sthlm |> 
  filter(!is.na(mht_status))
npx_sthlm_hrt <- npx_sthlm |> 
  filter(sample_id %in% clinical_sthlm_hrt$sample_id)

# Density
clinical_skane_density <- clinical_skane |> 
  filter(!is.na(stratus_densearea_cm2))
npx_skane_density <- npx_skane |> 
  filter(sample_id %in% clinical_skane_density$sample_id)

clinical_sthlm_density <- clinical_sthlm |> 
  filter(!is.na(stratus_densearea_cm2))
npx_sthlm_density <- npx_sthlm |> 
  filter(sample_id %in% clinical_sthlm_density$sample_id)
```

# Effect of clinical variables on NPX (e.g. age)

## Evaluating effect of plate_id

```{r}
plate_ev <- function(npx, clinical) {
  
  clinical <- clinical |> 
    select(sample_id, plate_id)
  
  npx_plate <- npx |> 
    left_join(clinical, by = "sample_id") |> 
    pivot_longer(cols = -c(sample_id, plate_id), names_to = "Protein", values_to = "value")  
  
  plot <- ggplot(npx_plate, aes(x = as.factor(plate_id), y = value)) +
    geom_boxplot(aes(fill = as.factor(plate_id))) +
    scale_fill_viridis_d(option = "magma") +
    theme_minimal() + 
    xlab("Plate ID") +
    ylab("NPX") +
    guides(fill = "none")
  
  return(plot)
}
```

```{r}
skane_plate <- plate_ev(npx_skane, clinical_skane)
sthlm_plate <- plate_ev(npx_sthlm, clinical_sthlm)
raw_plate <- plate_ev(npx_raw, sinfo_raw)

skane_plate + ggtitle("Skane") + sthlm_plate + ggtitle("Stockholm")
raw_plate + ggtitle("Raw data")
```

```{r}
skane_plate_ids <- clinical_skane |> 
  select(plate_id, sample_id)

sthlm_plate_ids <- clinical_sthlm |> 
  select(plate_id, sample_id)

common_proteins <- intersect(names(npx_skane), names(npx_sthlm)) 

common_skane <- npx_skane |> 
  select(all_of(common_proteins)) |> 
  mutate(Region = "Skane") |> 
  left_join(skane_plate_ids, by = "sample_id")
common_sthlm <- npx_sthlm |> 
  select(all_of(common_proteins)) |> 
  mutate(Region = "Stockholm") |> 
  left_join(sthlm_plate_ids, by = "sample_id")

combined_npx <- bind_rows(common_skane, common_sthlm) |> 
  pivot_longer(-c("sample_id", "plate_id", "Region"))
  
```

```{r}
combined_boxplots <- ggplot(combined_npx, aes(x = Region, y = value, fill = Region)) +
  geom_boxplot() +
  scale_fill_npg() +
  theme_minimal() + 
  xlab("Region") +
  ylab("NPX") +
  guides(fill = "none")

combined_boxplots
```

**Conclusion:** Plate ID differences within regions seem to be negligible. However, a slight difference is observed between regions. If the need to combine the data becomes relevant, then the region difference should be adjusted for. In other words, individual plate ids do not need to be adjusted for.

## Volcano plots, before/after NPX is adjusted for clinical variables

#### Age
```{r}
library(ggsci)
# Effect of age on NPX
skane_age <- DEA_func(npx_skane, clinical_skane, "age")
sthlm_age <- DEA_func(npx_sthlm, clinical_sthlm, "age")

skane_age_10 <- skane_age[1:10, ]
sthlm_age_10 <- sthlm_age[1:10, ]

age_skane <- volcano(skane_age) + ggtitle("Age", subtitle = "Skåne")
age_sthlm <- volcano(sthlm_age) + ggtitle(" ",subtitle = "Stockholm") + theme(axis.title.y = element_blank(), legend.position = "right")

age_skane + age_sthlm
```

#### BMI
```{r}
# Effect of bmi on NPX
skane_bmi <- DEA_func(npx_skane, clinical_skane, "bmi")
sthlm_bmi <- DEA_func(npx_sthlm, clinical_sthlm, "bmi")

skane_bmi_10 <- skane_bmi[1:10, ]
sthlm_bmi_10 <- sthlm_bmi[1:10, ]

bmi_skane <- volcano(skane_bmi) + ggtitle("BMI", subtitle = "Skåne") + ylim(0,30)
bmi_sthlm <- volcano(sthlm_bmi) + ggtitle(" ",subtitle = "Stockholm") + theme(axis.title.y = element_blank(), legend.position = "right") + ylim(0, 30)

bmi_skane + bmi_sthlm
```

#### Alcohol g/week
```{r}
# Effect of alcohol g/week on NPX
skane_alc <- DEA_func(npx_skane_alcohol, clinical_skane_alcohol, "alcohol_gram_week")
sthlm_alc<- DEA_func(npx_sthlm_alcohol, clinical_sthlm_alcohol, "alcohol_gram_week")

alc_skane <- volcano(skane_alc) + ggtitle("Alcohol (grams/week)", subtitle = "Skåne")
alc_sthlm <- volcano(sthlm_alc) + ggtitle(" ",subtitle = "Stockholm") + theme(axis.title.y = element_blank(), legend.position = "right")

alc_skane + alc_sthlm
```
some proteins are conserved between regions, some are not! useful to somewhat validate results
ano(sthlm_smoke) + ggtitle("Stockholm - Smoking status")

```{r}
# Removing effect of age before looking at density!

# Effect of density
skane_density <- DEA_func(npx_skane_density, clinical_skane_density, "stratus_densearea_cm2")
sthlm_density <- DEA_func(npx_sthlm_density, clinical_sthlm_density, "stratus_densearea_cm2")

density_skane <- volcano(skane_density) + ggtitle("Measured dense area of breast (cm²)", subtitle = "Skåne")
density_sthlm <- volcano(sthlm_density) + ggtitle(" ",subtitle = "Stockholm") + theme(axis.title.y = element_blank(), legend.position = "right")

density_skane + density_sthlm
```

```{r}
skane_sage <- DEA_func(npx_skane, clinical_skane, "sample_age")
sthlm_sage <- DEA_func(npx_sthlm, clinical_sthlm, "sample_age")

volcano(skane_sage) + ggtitle("Skane - sample age")
volcano(sthlm_sage) + ggtitle("Stockholm - sample age")
```

```{r}
skane_x <- DEA_func_logistic(npx_skane, clinical_skane, "x_BC") |> 
  mutate(logFC = Coefficient) 

sthlm_x <- DEA_func_logistic(npx_sthlm, clinical_sthlm, "x_BC") |> 
  mutate(logFC = Coefficient)

x_skane <- volcano_logistic(skane_x) + ggtitle("Breast cancer (case/control)", subtitle = "Skåne")
x_sthlm <- volcano_logistic(sthlm_x)+ ggtitle(" ",subtitle = "Stockholm") + theme(axis.title.y = element_blank(), legend.position = "right")

x_skane + x_sthlm
```

```{r}
skane_smoke <- DEA_func_logistic(npx_skane, clinical_skane, "smoking_status") |> 
  mutate(logFC = Coefficient)


sthlm_smoke <- DEA_func_logistic(npx_sthlm, clinical_sthlm, "smoking_status") |> 
  mutate(logFC = Coefficient) 

smoke_skane <- volcano_logistic(skane_smoke) + ggtitle("Smoking status (never, previous, current user)", subtitle = "Skåne")
smoke_sthlm <- volcano_logistic(sthlm_smoke)+ ggtitle(" ",subtitle = "Stockholm") + theme(axis.title.y = element_blank(), legend.position = "right")

smoke_skane + smoke_sthlm
```

```{r}
skane_mht <- DEA_func_logistic(npx_skane, clinical_skane, "mht_status") |> 
  mutate(logFC = Coefficient) 

sthlm_mht <- DEA_func_logistic(npx_sthlm, clinical_sthlm, "mht_status") |> 
  mutate(logFC = Coefficient) 

mht_skane <- volcano_logistic(skane_mht) + ggtitle("Hormonal treatment (never, previous, current user)", subtitle = "Skåne")
mht_sthlm <- volcano_logistic(sthlm_mht)+ ggtitle(" ",subtitle = "Stockholm") + theme(axis.title.y = element_blank(), legend.position = "right")

mht_skane + mht_sthlm
```
```{r}
clinical_skane <- clinical_skane |> 
  mutate(menopause = ifelse(menopause_status %in% c(1, 2), 1, 2))

clinical_sthlm <- clinical_sthlm |> 
  mutate(menopause = ifelse(menopause_status %in% c(1, 2), 1, 2))

skane_meno <- DEA_func_logistic(npx_skane, clinical_skane, "menopause") |> 
  mutate(logFC = Coefficient)

sthlm_meno <- DEA_func_logistic(npx_sthlm, clinical_sthlm, "menopause") |> 
  mutate(logFC = Coefficient) 

meno_skane <- volcano_logistic(skane_meno) + ggtitle("Menopause (pre- & peri vs. postmenopausal)", subtitle = "Skåne")
meno_sthlm <- volcano_logistic(sthlm_meno)+ ggtitle(" ",subtitle = "Stockholm") + theme(axis.title.y = element_blank(), legend.position = "right")

meno_skane + meno_sthlm
```

```{r}
skane_x <- DEA_func_logistic(npx_skane, clinical_skane |> filter(!(cancer_stage == 0)), "cancer_stage") |> 
  mutate(logFC = Coefficient)

volcano_logistic(skane_x) + ggtitle("Skane - cancer_stage")

sthlm_x <- DEA_func_logistic(npx_sthlm, clinical_sthlm |> filter(!(cancer_stage == 0)), "cancer_stage") |> 
  mutate(logFC = Coefficient)

volcano_logistic(sthlm_x) + ggtitle("Stockholm - cancer_stage")
```

```{r}
skane_x <- DEA_func_logistic(npx_skane, clinical_skane, "primary_blood_draw") |> 
  mutate(logFC = Coefficient)

volcano_logistic(skane_x) + ggtitle("Skane - primary_blood_draw")

sthlm_x <- DEA_func_logistic(npx_sthlm, clinical_sthlm, "primary_blood_draw") |> 
  mutate(logFC = Coefficient) 

volcano_logistic(sthlm_x) + ggtitle("Stockholm - primary_blood_draw")
```

```{r}
x_BC <- clinical_skane$x_BC
stratus_densearea_cm2 <- clinical_skane$stratus_densearea_cm2
  
model <- glm(x_BC ~ stratus_densearea_cm2, family = "binomial")
summary_model <- summary(model)

Coefficient <- summary_model$coefficients["stratus_densearea_cm2", "Estimate"]
PValue <- summary_model$coefficients["stratus_densearea_cm2", "Pr(>|z|)"]
AdjPValue <- p.adjust(PValue, method = "BH")
```

# Examining region differences
```{r}
protein_skane <- skane_age$protein
protein_sthlm <- sthlm_age$protein

# Find elements unique to df1
unique_to_skane <- setdiff(protein_skane, protein_sthlm)

# Find elements unique to df2
unique_to_sthlm <- setdiff(protein_sthlm, protein_skane)
```

CES1, IL6 both unique to sthlm, show up as top contenders in BMI

```{r}
# Need to remove proteins not present in both
common_proteins <- intersect(skane_age$protein, sthlm_age$protein)

skane_beta_age <- DEA_beta(npx_skane, clinical_skane, "age") |> 
  filter(protein %in% common_proteins) |> 
  rename("beta_skane" = label_data) |> 
  select(protein, beta_skane)

sthlm_beta_age <- DEA_beta(npx_sthlm, clinical_sthlm, "age") |> 
  filter(protein %in% common_proteins) |>
  rename("beta_sthlm" = label_data) |> 
  select(protein, beta_sthlm)

skane_beta_bmi <- DEA_beta(npx_skane, clinical_skane, "bmi") |> 
  filter(protein %in% common_proteins) |>
  rename("beta_skane" = label_data) |> 
  select(protein, beta_skane)

sthlm_beta_bmi <- DEA_beta(npx_sthlm, clinical_sthlm, "bmi") |> 
  filter(protein %in% common_proteins) |>
  rename("beta_sthlm" = label_data) |> 
  select(protein, beta_sthlm)

combined_beta_age <- skane_beta_age |> 
  left_join(sthlm_beta_age, by = "protein")

combined_beta_bmi <- skane_beta_bmi |> 
  left_join(sthlm_beta_bmi, by = "protein")
```

```{r}
# Same sign or not Skane/Sthlm
combined_beta_age$sign_group <- ifelse(sign(combined_beta_age$beta_skane) == sign(combined_beta_age$beta_sthlm), 'same_sign', 'diff_sign')
combined_beta_bmi$sign_group <- ifelse(sign(combined_beta_bmi$beta_skane) == sign(combined_beta_bmi$beta_sthlm), 'same_sign', 'diff_sign')

# In top 10 proteins for DE
combined_beta_age$label <- ifelse(combined_beta_age$protein %in% skane_age_10$protein | combined_beta_age$protein %in% sthlm_age_10$protein, combined_beta_age$protein, NA)
combined_beta_bmi$label <- ifelse(combined_beta_bmi$protein %in% skane_bmi_10$protein | combined_beta_bmi$protein %in% sthlm_bmi_10$protein, combined_beta_bmi$protein, NA)
```


```{r}
cor_age<- cor(combined_beta_age$beta_skane, combined_beta_age$beta_sthlm, use = "complete.obs")

ggplot(combined_beta_age, aes(x = beta_skane, y = beta_sthlm, color = sign_group)) +
  geom_point(alpha = 0.5) +
 # geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(x = "Skane", y = "Stockholm", title = expression("Comparing slopes ("*beta*") between regions"), subtitle = "Age") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_age, 3)), hjust = 3, vjust = 2) +
  scale_color_manual(values = c('same_sign' = '#4DBBD5FF', 'diff_sign' = '#E64B35FF'),
                     labels = c('Different sign (n = 14)', 'Same sign'),
                     name = "Slope direction") +
  theme_minimal()
```
```{r}
cor_bmi <- cor(combined_beta_bmi$beta_skane, combined_beta_bmi$beta_sthlm, use = "complete.obs")

ggplot(combined_beta_bmi, aes(x = beta_skane, y = beta_sthlm, color = sign_group)) +
  geom_point(alpha = 0.5) +
#  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(x = "Skane", y = "Stockholm", title = expression("Comparing slopes ("*beta*") between regions"), subtitle = "BMI") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_bmi, 3)), hjust = 2.3, vjust = 2) +
  scale_color_manual(values = c('same_sign' = '#4DBBD5FF', 'diff_sign' = '#E64B35FF'),
                     labels = c('Different sign (n = 24)', 'Same sign'),
                     name = "Slope direction") +
  theme_minimal()
  
```

```{r}
common_age_prot <- intersect(skane_age_10$protein, sthlm_age_10$protein) # 9/10 in common
common_bmi_prot <- intersect(skane_bmi_10$protein, sthlm_bmi_10$protein) # 5/10 in common

combined_common_bmi <- skane_bmi_10$protein |> 
  bind_cols(sthlm_bmi_10$protein)
counting_diff_sign_age <- combined_beta_age |> # 14
  filter(sign_group == "diff_sign")

counting_diff_sign_bmi <- combined_beta_bmi |> # 24
  filter(sign_group == "diff_sign")
```
## Distribution of clinical variables
```{r}
bmi_age_skane <- clinical_skane |> 
  select(sample_id, bmi, age) |> 
  mutate(Region = "Skane")

bmi_age_sthlm <- clinical_sthlm |> 
  select(sample_id, bmi, age) |> 
  mutate(Region = "Stockholm")

combined_bmi_age <- bind_rows(bmi_age_skane, bmi_age_sthlm)

plot_age_density <- ggplot(combined_bmi_age, aes(x = age, fill = Region)) +
  geom_density(alpha = 0.7) +
  labs(x = "Age", y = "Density", title = "Comparing age and BMI distributions between regions") +
  facet_wrap(~ Region) + 
  theme(legend.position = "none") + 
  scale_fill_npg() +
  theme_minimal()

plot_bmi_density <- ggplot(combined_bmi_age, aes(x = bmi, fill = Region)) +
  geom_density(alpha = 0.7) +
  labs(x = "BMI", y = "Density") +
  facet_wrap(~ Region) +
  theme(legend.position = "none") + 
  scale_fill_npg() +
  theme_minimal()

plot_age_density / plot_bmi_density
```
## Testing interaction term: Age + BMI in DEA
```{r}
DEA_interaction <- function(df, clin_df) {
  
  clin_df <- clin_df |> 
  select(age, bmi, sample_id)

  df <- df |>
  left_join(clin_df, by = "sample_id")
  
  label_data <- df |> 
    select(age, bmi)
  
  design_formula <- as.formula(paste("~", "age * bmi"))
  design_matrix <- model.matrix(design_formula, data = df)
  
  df <- df |> 
    select(-age, -bmi, -sample_id) |> 
    t()
  
  fit1 <- lmFit(df, design_matrix) # Linear fit
  fit2 <- eBayes(fit1) # Bayseian opt
  
  results <- topTable(fit2, adjust = "BH", number = Inf) |>  # Adjust multiple testing using Benjamini-Hochberg
    rownames_to_column(var = "protein")

  return(results)
}
```

```{r}
# Skane interaction
skane_interaction <- DEA_interaction(npx_skane, clinical_skane) |> 
  rename(logFC = age.bmi)
volcano(skane_interaction) + ggtitle("age + bmi", subtitle = "Skane")

# Stockholm interaction
sthlm_interaction <- DEA_interaction(npx_sthlm, clinical_sthlm) |> 
  rename(logFC = age.bmi) 
volcano(sthlm_interaction)+ ggtitle("age + bmi", subtitle = "Stockholm")
```
```{r}
# Association between age and bmi variables in both Skåne and Stockholm
# Skåne
predictor_skane <- clinical_skane[["age"]]
outcome_skane <- clinical_skane[["bmi"]]

age_bmi_skane <- glm(outcome_skane ~ predictor_skane)
summary_age_bmi <- summary(age_bmi_skane)

summary_age_bmi$coefficients

# Stockholm
predictor_sthlm <- clinical_sthlm[["age"]]

outcome_sthlm <- clinical_sthlm[["bmi"]]

age_bmi_sthlm <- glm(outcome_sthlm ~ predictor_sthlm)
summary_age_bmi <- summary(age_bmi_sthlm)

summary_age_bmi$coefficients
```


## Levene's test

```{r}
comb_regions <- bind_rows(clinical_skane, clinical_sthlm)

comb_regions$region <- as_factor(comb_regions$region)

levenes_age <- leveneTest(age ~ region, data = comb_regions, center = mean)
print(levenes_age)

levenes_bmi <- leveneTest(bmi ~ region, data = comb_regions, center = mean)
print(levenes_bmi)

```

```{r}
# dense area
levenes_dense <- leveneTest(stratus_densearea_cm2 ~ region, data = comb_regions, center = mean)
print("dense area")
print(levenes_dense)


# menopause
levenes_meno <- leveneTest(menopause_status ~ region, data = comb_regions, center = mean)
print("meno")
print(levenes_meno)

# hormonal
levenes_mht <- leveneTest(mht_status ~ region, data = comb_regions, center = mean)
print("hormonal")
print(levenes_mht)

# alcohol
levenes_alc <- leveneTest(alcohol_gram_week ~ region, data = comb_regions, center = mean)
print("alc")
print(levenes_alc)

# smoking
levenes_smoke <- leveneTest(smoking_status ~ region, data = comb_regions, center = mean)
print("smoke")
print(levenes_smoke)
```


```{r}
combined_age_binned <- combined_bmi_age |>
  mutate(age_group = case_when(
    age < 50 ~ "<50 (n = 383)",               
    age >= 50 & age <= 65 ~ "50-65 (n = 804)",  
    TRUE ~ ">65 (n = 469)"                    
  ))

combined_age_binned$age_group <- as_factor(combined_age_binned$age_group)
```

```{r}
levenes_age_bmi <- leveneTest(bmi ~ age_group, combined_age_binned, center = mean)
print(levenes_age_bmi)
```
```{r}
levenes_skane <- combined_age_binned |> 
  filter(Region == "Skane")

leveneTest(bmi ~ age_group, levenes_skane, center = mean)

levenes_sthlm <- combined_age_binned |> 
  filter(Region == "Stockholm")

leveneTest(bmi ~ age_group, levenes_sthlm, center = mean)
```


```{r}
age_group_count <- combined_age_binned |> 
  group_by(age_group) |> 
  summarise(count = n())
```


```{r}
plot_age_binned_density <- ggplot(combined_age_binned, aes(x = age, fill = Region)) +
  geom_density(alpha = 0.7) +
  labs(x = "Age", y = "Density", title = "Age distribution between regions") +
  facet_wrap(~ Region) + 
  theme(legend.position = "none") +
  geom_vline(xintercept = 50, linetype = "dashed", alpha = 0.7) +
  geom_vline(xintercept = 65, linetype = "dashed", alpha = 0.7) +
  theme_minimal() +
  scale_fill_npg()

plot_age_bmi <- ggplot(combined_age_binned, aes(x = age, y = bmi, fill = age_group)) +
  geom_boxplot(alpha = 0.7, width = 0.1, varwidth = TRUE) +
 # scale_fill_npg() +
  labs(fill = "Age group") +
  theme_minimal()

plot_age_binned_density
plot_age_bmi + ggtitle("BMI across age groups")
```

```{r}
adjuster_interaction <- function(df, clin_df) {
  
  # Ensure 'age' and 'bmi' are selected along with 'sample_id'
  clin_df <- clin_df |> 
    select(age, bmi, sample_id)
  
  # Merge the clinical columns into the protein expression dataframe
  df <- df |>
    left_join(clin_df, by = "sample_id")
  
  # Store sample ids
  sample_ids <- df$sample_id
  
  # Retrieve column names for protein data, excluding 'age', 'bmi', and 'sample_id'
  col_names <- setdiff(colnames(df), c("age", "bmi", "sample_id"))
   
  # Initialize a tibble to store adjusted protein levels
  adjusted_npx <- tibble(matrix(ncol = length(col_names), nrow = nrow(df)))
  names(adjusted_npx) <- col_names  # Set correct column names
  
  # Loop through each protein column to calculate adjusted values
  for (protein in col_names) {
    
    # Fitting a linear model with interaction between age and bmi
    linear_model <- lm(reformulate(termlabels = c("age", "bmi", "age:bmi"), response = protein), data = df)
    
    # Storing the residuals, which represent protein levels adjusted for age, bmi, and their interaction
    adjusted_npx[[protein]] <- linear_model$residuals
  }

  # Combine the sample_id back with the adjusted protein data
  adjusted_npx <- bind_cols(tibble(sample_id = sample_ids), adjusted_npx)
  
  # Return the adjusted protein expression data
  return(adjusted_npx)
}
```

```{r}
# Removing effect of interaction term of age:bmi
adjusted_skane <- adjuster_interaction(npx_skane, clinical_skane)
adjusted_sthlm <- adjuster_interaction(npx_sthlm, clinical_sthlm)
```

```{r}
data <- data.frame(
  Formula = c("age ~ region", "bmi ~ region", "breast dense area ~ region","alcohol g/week ~ region", "hormonal_treatment ~ region", "bmi ~ age_group", "bmi ~ age_group", "bmi ~ age_group"),
  P_Value = c(0.9123, 0.5928, 0.01269, 3.275e-05, 0.01975, 0.001022, 0.001327, 0.1382),
  Region = c("Skåne + Stockholm", "Skåne + Stockholm", "Skåne + Stockholm", "Skåne + Stockholm", "Skåne + Stockholm", "Skåne + Stockholm", "Skåne", "Stockholm")
)

# Function to render the table using gt
render_table <- function(data) {
  # Render the table with gt
  data %>%
    gt() %>%
    cols_width(
      Formula ~ px(300),  # Specify width in pixels using formula
      P_Value ~ px(100)
    ) %>%
    cols_label(
      Formula = "Formula",
      P_Value = "P-Value"
    ) %>%
    fmt_number(
      columns = c(P_Value),
      decimals = 5
    ) %>%
    tab_header(
      title = "Levene's Test Results"
    ) %>%
    tab_style(
      style = cell_text(weight = "bold"),  # Apply bold style to column labels
      locations = cells_column_labels(columns = everything())
    ) %>%
    tab_style(
      style = cell_text(style = "italic"),  # Make text bold, italic, and red
      locations = cells_body(
        columns = P_Value,
        rows = P_Value < 0.05  # Apply this style to rows where P_Value is less than 0.05
      )
    )
}

# Create and display the table
render_table(data)
```


## Output
```{r}
# Output from adjusting for variables
write_csv(adjusted_skane, "../clean_data/npx_skane_adj.csv")
write_csv(adjusted_sthlm, "../clean_data/npx_sthlm_adj.csv")
```


```{r}

```

