# Effects of AbsPQN analysis

Loading packages and data.

```{r}
# Packages
library(tidyverse)
library(viridis)
library(pheatmap)
library(umap)
library(corrplot)
library(patchwork)

# Clinical data
clinical_skane <- read_csv("../clean_data/clinical_skane_filtered.csv", show_col_types = FALSE)
clinical_sthlm <- read_csv("../clean_data/clinical_sthlm_filtered.csv", show_col_types = FALSE)
```

Here, a function is created which will load the specified NPX data frame and also clean/sort it.

```{r}
# Function for cleaning and sorting the input data frames
cleaner <- function(path, clinical_df) {
  
  clean_df <- read_csv(path, show_col_types = FALSE) |> # Loads the data
    left_join(clinical_df |> select(sample_id, individual_id), by = "sample_id") |> # Attaches sample- and individual id to NPX data
    select(sample_id, individual_id, everything()) |> # Reorders columns to put sample- and individual id at front
    arrange(individual_id) # Sorts the rows by their individual id
  
  return(clean_df)
}
```

Calling the cleaner function..

```{r}
## Pre AbsPQN
pre_skane <- cleaner("../clean_data/npx_skane_filtered.csv", clinical_skane)
pre_sthlm <- cleaner("../clean_data/npx_sthlm_filtered.csv", clinical_sthlm)

## Post AbsPQN

# AbsPQN1
post_skane_v1 <- cleaner("../clean_data/norm_npx_skane_v1.csv", clinical_skane)
post_sthlm_v1 <- cleaner("../clean_data/norm_npx_sthlm_v1.csv", clinical_sthlm)

# AbsPQN2
post_skane_v2 <- cleaner("../clean_data/norm_npx_skane_v2.csv", clinical_skane)
post_sthlm_v2 <- cleaner("../clean_data/norm_npx_sthlm_v2.csv", clinical_sthlm)
```

# Correlations

This section will examine the effect of what format the data has when AbsPQN is applied (log2 or not).

## Protein-protein correlations

Here, protein-protein correlations are calculated and then compared before and after AbsPQN.

```{r}
# Protein-protein correlations function
protein_corr <- function(df, df_name) {
 
  # Creates correlation matrix
   corr_df <- df |> 
    select(-sample_id, -individual_id) |> 
    cor(method = "spearman", use = "complete.obs") 
  
  # Removes diagonal
  diag(corr_df) <- NA
  
  # Long format
  long_corr_df <- corr_df |> 
    as_tibble(corr_df, rownames = "Protein1") |> 
    pivot_longer(-Protein1, names_to = "Protein2", values_to = "Correlation") |> 
    mutate(Matrix = df_name)
  
  return(long_corr_df)
}
```

Calling correlation function..
```{r}
## Pre AbsPQN
pre_skane_corr <- protein_corr(pre_skane, "pre_skane_NA")
pre_sthlm_corr <- protein_corr(pre_sthlm, "pre_sthlm_NA")

## Post AbsPQN

# AbsPQN1
post_skane_corr_v1 <- protein_corr(post_skane_v1, "post_skane_v1")
post_sthlm_corr_v1 <- protein_corr(post_sthlm_v1, "post_sthlm_v1")

# AbsPQN2
post_skane_corr_v2 <- protein_corr(post_skane_v2, "post_skane_v2")
post_sthlm_corr_v2 <- protein_corr(post_sthlm_v2, "post_sthlm_v2")

# Combining all of the data
combined_corr <- bind_rows(pre_skane_corr, pre_sthlm_corr, post_skane_corr_v1, post_sthlm_corr_v1, post_skane_corr_v2, post_sthlm_corr_v2) |> 
  separate_wider_delim(delim = "_", cols = Matrix, names = c("AbsPQN", "Region", "v1/v2"))
```

Creating plots, one for AbsPQN1 and one for AbsPQN 2 to allow for comparison of result.

```{r}
# AbsPQN1
a <- combined_corr |> 
  filter(`v1/v2` != "v2") |> 
  ggplot(aes(x = Correlation, fill = AbsPQN, group = AbsPQN)) +
    geom_density(alpha = 0.5) +
    facet_wrap(~ Region, ncol = 2) +
    labs(title = "Protein-Protein correlation, pre/post AbsPQN", subtitle = "AbsPQN1", x = "Correlation", y = "Density") +
    theme_minimal() +
    scale_x_continuous(limits = c(-1, 1)) +
    scale_y_continuous(limits = c(0, 4)) +
    scale_fill_npg() +
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3) +
    geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.3) +
    geom_vline(xintercept = -0.5, linetype = "dashed", alpha = 0.3)

# AbsPQN2
b <- combined_corr |> 
  filter(`v1/v2` != "v1") |> 
  ggplot(aes(x = Correlation, fill = AbsPQN, group = AbsPQN)) +
    geom_density(alpha = 0.5) +
    facet_wrap(~ Region, ncol = 2) +
    labs(title = "Protein-Protein correlation", subtitle = "AbsPQN2", x = "Correlation", y = "Density") +
    theme_minimal() +
    scale_x_continuous(limits = c(-1, 1)) +
    scale_y_continuous(limits = c(0, 4)) +
    scale_fill_npg() +
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3) +
    geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.3) +
    geom_vline(xintercept = -0.5, linetype = "dashed", alpha = 0.3)
```

Plotting..

```{r}
a 
b
```

Aside from density plot, another way to visualize the change between pre/post AbsPQN data is to plot them against each other in a scatter plot.The diagonal of the plot represents perfect correlation between pre/post.

```{r}
# Creating new column with protein pairs instead of separate proteins
combined_corr <- combined_corr |> 
  mutate(Corr_pairs = paste(Protein1, Protein2, sep = ","))
```

The following function will plot the scatter plots for both AbsPQN v1 and v2, and whether it comes from Skane or Stockholm.

```{r}
# Function for creating the scatterplots
pre_post_plot <- function(df, method, region) {
  
  # Changes output, depending on whether data is AbsPQN 1 or 2
  if (method == "v1") {
    df <- filter(df, `v1/v2` != "v2")
    method_title <- "AbsPQN1"
  } else {
    df <- filter(df, `v1/v2` != "v1")
    method_title <- "AbsPQN2"
  }
  
  # Changes output depending on the region
  if (region == "skane") {
    filter(df, Region == "skane")
    region_title <- "Skane"
    
  } else {
    filter(df, Region == "sthlm")
    region_title <- "Sthlm"
  }
  
  # Split into pre and post
  pre <- filter(df, AbsPQN == "pre") |> 
    select(Corr_pairs, Correlation)
  post <- filter(df, AbsPQN == "post") |> 
    select(Corr_pairs, Correlation)
  
  # Merge on "Corr_pairs"
  combined <- merge(pre, post, by = "Corr_pairs", suffixes = c("_pre", "_post"))
  
  plot <- ggplot(combined, aes(x = Correlation_pre, y = Correlation_post)) +
    geom_point(color = "#4DBBD5FF", alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
    labs(x = "Pre AbsPQN", y = "Post AbsPQN", title = region_title, subtitle = method_title) +
    theme_minimal()

  return(plot)
}
```

Calling the scatter plot function..

```{r}
skane_v1_plot <- pre_post_plot(combined_corr, "v1", "skane")
skane_v2_plot <- pre_post_plot(combined_corr, "v2", "skane")

sthlm_v1_plot <- pre_post_plot(combined_corr, "v1", "sthlm")
sthlm_v2_plot <- pre_post_plot(combined_corr, "v2", "sthlm")
```

Plotting..

```{r}
# Skane
skane_v1_plot
skane_v2_plot
```

```{r}
# Stockholm
sthlm_v1_plot
sthlm_v2_plot
```

# Comparing primary & secondary blood draw

First, we create a function which will extract the primary/secondary blood draw samples from the incoming NPX data frames and calculate the sample-sample correlations.
```{r}
prim_sec_corr <- function(clinical_df, npx_df, pre_post, method) {
  # Filtering clinical data for samples with primary/secondary blood draws
  sec <- clinical_df |> 
    filter(primary_blood_draw == 0)
  
  prim <- clinical_df |> 
    filter(individual_id %in% sec$individual_id) |> 
    filter(primary_blood_draw == 1) 

  # Filtering NPX data frame to match the clinical data
  sec_npx <- npx_df |> 
    filter(sample_id %in% sec$sample_id) |>
    arrange(individual_id) |> # Ensures that the orders match between primary and secondary
    select(-sample_id, -individual_id) |> # Removing both IDs so that they are not included in correlation calculations
    t() # Transposes the data frame since we want to explore sample-sample correlation
  
  prim_npx <- npx_df |> 
    filter(sample_id %in% prim$sample_id) |> 
    arrange(individual_id) |> 
    select(-sample_id, -individual_id) |> 
    t()

  # Calculating correlation between each primary/secondary sample
  correlations <- numeric(ncol(prim_npx)) # Empty data frame to hold correlation values

  for(i in 1:ncol(prim_npx)) {
    correlations[i] <- cor(prim_npx[, i], sec_npx[, i], use = "complete.obs")
  }
  
  colnames(correlations) <- colnames(prim_npx)
  
  correlations <- as_tibble(correlations) |> 
    mutate(`Pre/Post` = pre_post) |> 
    mutate(`v1/v2` = method)
    
  return(correlations)
}
```
Calling the function and calculating the correlations. Also combining the data frames.
```{r}
# Pre AbsPQN
pre_skane_prim_sec <- prim_sec_corr(clinical_skane, pre_skane, "pre", NA)
pre_sthlm_prim_sec <- prim_sec_corr(clinical_sthlm, pre_sthlm, "pre", NA)

# Post AbsPQN (v1 or v2)
post_v1_skane_prim_sec <- prim_sec_corr(clinical_skane, post_skane_v1, "post", "v1")
post_v2_skane_prim_sec <- prim_sec_corr(clinical_skane, post_skane_v2, "post", "v2")

post_v1_sthlm_prim_sec <- prim_sec_corr(clinical_sthlm, post_sthlm_v1, "post", "v1")
post_v2_sthlm_prim_sec <- prim_sec_corr(clinical_sthlm, post_sthlm_v2, "post", "v2")

# Combining per region
combined_skane <- bind_rows(pre_skane_prim_sec, post_v1_skane_prim_sec, post_v2_skane_prim_sec)

combined_sthlm <- bind_rows(pre_sthlm_prim_sec, post_v1_sthlm_prim_sec, post_v2_sthlm_prim_sec)

```

Now, we need to create the actual density plots.
```{r}

ggplot(combined_skane, aes(x = value, y = after_stat(scaled), fill = `Pre/Post`, group = `Pre/Post`)) +
  geom_density( alpha = 0.5) +
  facet_wrap(~ `v1/v2`, ncol = 1) +
  labs(x = "Correlation", y = "Density", title = "Comparing correlation primary/secondary", subtitle = "Skane") +
  theme_minimal() +
  scale_x_continuous(limits = c(0.8, 1)) +
  scale_fill_npg() +
  theme(legend.position = "top") +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = -0.5, linetype = "dashed", alpha = 0.3)

ggplot(combined_sthlm, aes(x = value, y = after_stat(scaled), fill = `Pre/Post`, group = `Pre/Post`)) +
  geom_density( alpha = 0.5) +
  facet_wrap(~ `v1/v2`, ncol = 1) +
  labs(x = "Correlation", y = "Density", title = "Comparing correlation primary/secondary", subtitle = "Sthlm") +
  theme_minimal() +
  scale_x_continuous(limits = c(0.8, 1)) +
  scale_fill_npg() +
  theme(legend.position = "top") +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = -0.5, linetype = "dashed", alpha = 0.3)
```
To summarize the differences between data before and after AbsPQN, a density plot for each region which contains the correlation between individual samples/protein before and after AbsPQN.
```{r}
# Quick correlation function
correlater <- function(pre_df, post_df, method = "default") {
  
  filt_pre_df <- pre_df |> 
    select(-sample_id, -individual_id)
  filt_sec_df <- post_df |> 
    select(-sample_id, -individual_id)
  
  if (method == "sample") {
    filt_pre_df <- t(filt_pre_df)
    filt_sec_df <- t(filt_sec_df)
  } else {
    
  }
  
  correlations <- numeric(ncol(filt_pre_df)) # Empty data frame to hold correlation values

  for(i in 1:ncol(filt_pre_df)) {
    correlations[i] <- cor(filt_pre_df[, i], filt_sec_df[, i], use = "complete.obs")
  }
  
  correlations_df <- tibble(correlation = correlations)
    
  return(correlations_df)
}

# Sample-sample correlation before and after AbsPQN (i.e. comparing Sample ID = 1 in pre_skane with Sample ID = 1 in post_skane)

sample_skane_corr_v1 <- correlater(pre_skane, post_skane_v1, "sample") |> 
  mutate(Method = "v1")
sample_sthlm_corr_v1 <- correlater(pre_sthlm, post_sthlm_v1, "sample") |> 
  mutate(Method = "v1")

sample_skane_corr_v2 <- correlater(pre_skane, post_skane_v2, "sample") |> 
  mutate(Method = "v2")
sample_sthlm_corr_v2 <- correlater(pre_sthlm, post_sthlm_v2, "sample") |> 
  mutate(Method = "v2")

combined_sample_skane_corr <- bind_rows(sample_skane_corr_v1, sample_skane_corr_v2) |> 
  mutate(Type = "sample-sample")

combined_sample_sthlm_corr <- bind_rows(sample_sthlm_corr_v1, sample_sthlm_corr_v2) |> 
  mutate(Type = "sample-sample")

# Protein-protein correlation 
prot_skane_corr_v1 <- correlater(pre_skane, post_skane_v1) |> 
  mutate(Method = "v1")
prot_sthlm_corr_v1 <- correlater(pre_sthlm, post_sthlm_v1) |> 
  mutate(Method = "v1")

prot_skane_corr_v2 <- correlater(pre_skane, post_skane_v2) |> 
  mutate(Method = "v2")
prot_sthlm_corr_v2 <- correlater(pre_sthlm, post_sthlm_v2) |> 
  mutate(Method = "v2")

combined_prot_skane_corr <- bind_rows(prot_skane_corr_v1, prot_skane_corr_v2) |> 
  mutate(Type = "protein-protein")

combined_prot_sthlm_corr <- bind_rows(prot_sthlm_corr_v1, prot_sthlm_corr_v2) |> 
  mutate(Type = "protein-protein")

# Final combination before plotting
combined_corr_skane <- bind_rows(combined_sample_skane_corr, combined_prot_skane_corr)
combined_corr_sthlm <- bind_rows(combined_sample_sthlm_corr, combined_prot_sthlm_corr)
```

Creating density plots:
```{r}
ggplot(combined_corr_skane, aes(x = correlation, y = after_stat(scaled), fill = Type, group = Type)) +
  geom_density(alpha = 0.8) +
  facet_wrap(~ Method, ncol = 2) +
  labs(x = "Correlation", y = "Density", title = "Correlation sample-sample and protein-protein", subtitle = "Skane") +
  theme_minimal() +
  scale_fill_npg() +
  scale_x_continuous(limits = c(0, 1)) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = -0.5, linetype = "dashed", alpha = 0.3)

ggplot(combined_corr_sthlm, aes(x = correlation, y = after_stat(scaled), fill = Type, group = Type)) +
  geom_density(alpha = 0.8) +
  facet_wrap(~ Method, ncol = 2) +
  labs(x = "Correlation", y = "Density", title = "Correlation sample-sample and protein-protein", subtitle = "Stockholm") +
  theme_minimal() +
  scale_fill_npg() +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = -0.5, linetype = "dashed", alpha = 0.3)
```
# Plotting pre-npx against post-npx, scatter plot
kaos
```{r}
# Skane
long_skane_pre <- pre_skane |> 
  arrange(sample_id) |> 
  select(-sample_id, -individual_id) |> 
  pivot_longer(everything()) |> 
  rename(value_pre = value)

long_skane_post_v1 <- post_skane_v1 |> 
  arrange(sample_id) |> 
  select(-sample_id, -individual_id) |> 
  pivot_longer(everything()) |> 
  rename(value_post = value)

long_skane_post_v2 <- post_skane_v2 |> 
  arrange(sample_id) |> 
  select(-sample_id, -individual_id) |> 
  pivot_longer(everything()) |> 
  rename(value_post = value)

v1_scatter_skane <- bind_cols(long_skane_pre, long_skane_post_v1$value_post) |> 
  rename(value_post = ...3)
v2_scatter_skane <- bind_cols(long_skane_pre, long_skane_post_v2$value_post) |> 
  rename(value_post = ...3)
```
```{r}
plot_skane_1 <- ggplot(v1_scatter_skane, aes(x = value_pre, y = value_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = "Pre AbsPQN", y = "Post AbsPQN",  title = "NPX for each protein pre/post", subtitle = "Skane - AbsPQN1") +
  theme_minimal()

plot_skane_2 <- ggplot(v2_scatter_skane, aes(x = value_pre, y = value_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = "Pre AbsPQN", y = "Post AbsPQN", title = "NPX for each protein pre/post", subtitle = "Skane - AbsPQN2") +
  theme_minimal()
```


```{r}
# Stockholm
long_sthlm_pre <- pre_sthlm |> 
  arrange(sample_id) |> 
  select(-sample_id, -individual_id) |> 
  pivot_longer(everything()) |> 
  rename(value_pre = value)

long_sthlm_post_v1 <- post_sthlm_v1 |> 
  arrange(sample_id) |> 
  select(-sample_id, -individual_id) |> 
  pivot_longer(everything()) |> 
  rename(value_post = value)

long_sthlm_post_v2 <- post_sthlm_v2 |> 
  arrange(sample_id) |> 
  select(-sample_id, -individual_id) |> 
  pivot_longer(everything()) |> 
  rename(value_post = value)

v1_scatter_sthlm <- bind_cols(long_sthlm_pre, long_sthlm_post_v1$value_post) |> 
  rename(value_post = ...3)
v2_scatter_sthlm <- bind_cols(long_sthlm_pre, long_sthlm_post_v2$value_post) |> 
  rename(value_post = ...3)


```

```{r}
plot_sthlm_1 <- ggplot(v1_scatter_sthlm, aes(x = value_pre, y = value_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = "Pre AbsPQN", y = "Post AbsPQN", subtitle = "Sthlm - AbsPQN1") +
  theme_minimal()

plot_sthlm_2 <- ggplot(v2_scatter_sthlm, aes(x = value_pre, y = value_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = "Pre AbsPQN", y = "Post AbsPQN", subtitle = "Sthlm - AbsPQN2") +
  theme_minimal()
```

```{r}
plot_skane_1 + plot_skane_2
  
  plot_sthlm_1+ plot_sthlm_2
```

# Plotting protein before and after correlation, per protein

Here we plot the NPX before against after AbsPQN, for each protein in a separate scatter plot. All of the plots are saved in a PDF.
```{r}
# Pre AbsPQN
pre_skane_long <- pre_skane |> 
    pivot_longer(
    -sample_id, 
    names_to = "protein", 
    values_to = "value_pre"
  )
pre_sthlm_long <- pre_sthlm |> 
    pivot_longer(
    -sample_id, 
    names_to = "protein", 
    values_to = "value_pre"
  )


# Post AbsPQN
post_skane_long <- post_skane_v2 |> 
    pivot_longer(
    -sample_id, 
    names_to = "protein", 
    values_to = "value_post"
  )
post_sthlm_long <- post_sthlm_v2 |> 
    pivot_longer(
    -sample_id, 
    names_to = "protein", 
    values_to = "value_post"
  )
```


```{r}
# Skane PDF
skane_combined <- merge(pre_skane_long, post_skane_long, by = c("sample_id", "protein"), suffixes = c("_pre", "_post"))

correlation_df <- skane_combined |> 
  group_by(protein) |> 
  summarize(correlation = cor(value_pre, value_post, use = "complete.obs")) |> 
  ungroup()

skane_combined_with_corr <- merge(skane_combined, correlation_df, by = "protein")

# Adjust these variables as needed
facets_per_page <- 20  # Number of plots per page

total_proteins <- length(unique(skane_combined_with_corr$protein))
total_pages <- ceiling(total_proteins / facets_per_page)

plot_page <- function(page_number, facets_per_page, data) {
  start <- ((page_number - 1) * facets_per_page) + 1
  end <- min(start + facets_per_page - 1, length(unique(data$protein)))
  
  protein_names <- unique(data$protein)[start:end]
  data_subset <- data[data$protein %in% protein_names,]
  
  data_subset$protein_corr_label <- with(data_subset, paste(protein, sprintf("(r = %.2f)", correlation), sep=" "))
  
  plots <- ggplot(data_subset, aes(x = value_pre, y = value_post)) + 
    geom_point(color = "#4DBBD5FF", alpha = 0.5) +
    facet_wrap(~ protein_corr_label, scales = "free", ncol = 4) +  # Adjust ncol as needed
    theme_minimal() +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#265073") +
    xlab("Pre Expression") +
    ylab("Post Expression") +
    ggtitle(sprintf("Protein Expression Pre vs. Post AbsPQN2 (Page %d)", page_number))
  
  return(plots)
}

# Calling
pdf("skane_pre_post_abspqn_per_protein.pdf", width = 11, height = 8.5)
for (page in 1:total_pages) {
  print(plot_page(page, facets_per_page, skane_combined_with_corr))
}
dev.off()
```


```{r}
# Stockholm PDF
sthlm_combined <- merge(pre_sthlm_long, post_sthlm_long, by = c("sample_id", "protein"), suffixes = c("_pre", "_post"))

correlation_df <- sthlm_combined |> 
  group_by(protein) |> 
  summarize(correlation = cor(value_pre, value_post, use = "complete.obs")) |> 
  ungroup()

sthlm_combined_with_corr <- merge(sthlm_combined, correlation_df, by = "protein")

total_proteins <- length(unique(sthlm_combined_with_corr$protein))
total_pages <- ceiling(total_proteins / facets_per_page)

pdf("sthlm_pre_post_abspqn_per_protein.pdf", width = 11, height = 8.5)
for (page in 1:total_pages) {
  print(plot_page(page, facets_per_page, sthlm_combined_with_corr))
}
dev.off()
```


## Comparing IQR, Sample Median before and after AbsPQN

```{r}
median_iqr <- function(df) {
  
  df <- df |> 
    select(-sample_id, -individual_id)
  
  medians <- apply(df, 1, median)
  
  # Compute IQR for each row
  iqrs <- apply(df, 1, IQR)
  
  # Combine medians and IQRs into a results dataframe
  results <- data.frame(median = medians, iqr = iqrs)

  return(results)
}

# Pre
median_iqr_skane <- median_iqr(pre_skane) |> 
  rename(median_pre = median) |> 
  rename(iqr_pre = iqr)
median_iqr_sthlm <- median_iqr(pre_sthlm) |> 
  rename(median_pre = median) |> 
  rename(iqr_pre = iqr)

# AbsPQN1
median_iqr_skane_v1 <- median_iqr(post_skane_v1)|> 
  rename(median_post = median) |> 
  rename(iqr_post = iqr)
median_iqr_sthlm_v1 <- median_iqr(post_sthlm_v1) |> 
  rename(median_post = median) |> 
  rename(iqr_post = iqr)

# AbsPQN2
median_iqr_skane_v2 <- median_iqr(post_skane_v2)|> 
  rename(median_post = median) |> 
  rename(iqr_post = iqr)
median_iqr_sthlm_v2 <- median_iqr(post_sthlm_v2) |> 
  rename(median_post = median) |> 
  rename(iqr_post = iqr)


# Combining into pairs

# Skane
skane_median_iqr_v1 <- bind_cols(median_iqr_skane, median_iqr_skane_v1)
skane_median_iqr_v2 <- bind_cols(median_iqr_skane, median_iqr_skane_v2)

# Sthlm
sthlm_median_iqr_v1 <- bind_cols(median_iqr_sthlm, median_iqr_sthlm_v1)
sthlm_median_iqr_v2 <- bind_cols(median_iqr_sthlm, median_iqr_sthlm_v2)


```

```{r}
# IQR comparison, SKANE
cor_iqr_skane_v1 <- cor(skane_median_iqr_v1$iqr_pre, skane_median_iqr_v1$iqr_post, use = "complete.obs")

iqr_skane_1 <- ggplot(skane_median_iqr_v1, aes(x = iqr_pre, y = iqr_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " IQR - Pre AbsPQN", y = "IQR - Post AbsPQN",  title = "IQR", subtitle = "Skane - AbsPQN1") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_iqr_skane_v1, 3)), hjust = 3, vjust = 2) +
  theme_minimal()

cor_iqr_skane_v2 <- cor(skane_median_iqr_v2$iqr_pre, skane_median_iqr_v2$iqr_post, use = "complete.obs")

iqr_skane_2 <- ggplot(skane_median_iqr_v2, aes(x = iqr_pre, y = iqr_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " IQR - Pre AbsPQN", y = "IQR - Post AbsPQN",  title = "IQR", subtitle = "Skane - AbsPQN2") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_iqr_skane_v2, 3)), hjust = 3, vjust = 2) +
  theme_minimal()

# IQR comparison, STHLM
cor_iqr_sthlm_v1 <- cor(sthlm_median_iqr_v1$iqr_pre, sthlm_median_iqr_v1$iqr_post, use = "complete.obs")

iqr_sthlm_1 <- ggplot(sthlm_median_iqr_v1, aes(x = iqr_pre, y = iqr_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " IQR - Pre AbsPQN", y = "IQR - Post AbsPQN",  title = "IQR", subtitle = "Sthlm - AbsPQN1") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_iqr_sthlm_v1, 3)), hjust = 3, vjust = 2) +
  theme_minimal()

cor_iqr_sthlm_v2 <- cor(sthlm_median_iqr_v2$iqr_pre, sthlm_median_iqr_v2$iqr_post, use = "complete.obs")

iqr_sthlm_2 <- ggplot(sthlm_median_iqr_v2, aes(x = iqr_pre, y = iqr_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " IQR - Pre AbsPQN", y = "IQR - Post AbsPQN",  title = "IQR", subtitle = "Sthlm - AbsPQN2") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_iqr_sthlm_v2, 3)), hjust = 3, vjust = 2) +
  theme_minimal()
```

```{r}
# Calling IQR plots

iqr_skane_1 + iqr_sthlm_1

iqr_skane_2 + iqr_sthlm_2
```


```{r}
# Median comparison, SKANE
cor_skane_v1 <- cor(skane_median_iqr_v1$median_pre, skane_median_iqr_v1$median_post, use = "complete.obs")


median_skane_1 <- ggplot(skane_median_iqr_v1, aes(x = median_pre, y = median_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " Median - Pre AbsPQN", y = "IQR - Post AbsPQN",  title = "Median", subtitle = "Skane - AbsPQN1") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_skane_v1, 3)), hjust = 3, vjust = 2) +
  theme_minimal()

cor_skane_v2 <- cor(skane_median_iqr_v2$median_pre, skane_median_iqr_v2$median_post, use = "complete.obs")

median_skane_2 <- ggplot(skane_median_iqr_v2, aes(x = median_pre, y = median_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " Median - Pre AbsPQN", y = "IQR - Post AbsPQN",  title = "Median", subtitle = "Skane - AbsPQN2") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_skane_v2, 3)), hjust = 3, vjust = 2) +
  theme_minimal()

# Median comparison, STHLM
cor_sthlm_v1 <- cor(sthlm_median_iqr_v1$median_pre, sthlm_median_iqr_v1$median_post, use = "complete.obs")

median_sthlm_1 <- ggplot(sthlm_median_iqr_v1, aes(x = median_pre, y = median_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " Median - Pre AbsPQN", y = "IQR - Post AbsPQN",  title = "Median", subtitle = "Sthlm - AbsPQN1") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_sthlm_v1, 3)), hjust = 3, vjust = 2) +
  theme_minimal()

cor_sthlm_v2 <- cor(sthlm_median_iqr_v2$median_pre, sthlm_median_iqr_v2$median_post, use = "complete.obs")

median_sthlm_2 <- ggplot(sthlm_median_iqr_v2, aes(x = median_pre, y = median_post)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " Median - Pre AbsPQN", y = "IQR - Post AbsPQN",  title = "Median", subtitle = "Sthlm - AbsPQN2") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_skane_v1, 3)), hjust = 3, vjust = 2) +
  theme_minimal()
```

```{r}
# Calling median plots
median_skane_1 + median_sthlm_1

median_skane_2 + median_sthlm_2
```

### Does AbsPQN2 increase similarity between Skane and Stockholm?
Plotting IQRs against each other to see how their relationship changes after AbsPQN2.

IQR FOR SAMPLES


The code is very chaotic from here on out...


```{r}
# Since we compare between regions, need to combine data sets differently compared to how it was done above

region_iqr_skane <- median_iqr_skane |> 
  select(iqr_pre) |> 
  rename(skane_iqr = iqr_pre)

region_iqr_sthlm <- median_iqr_sthlm |> 
  select(iqr_pre) |> 
  rename(sthlm_iqr = iqr_pre)

# Different length Skane and Sthlm, trim some elements from the longer df
region_iqr_sthlm_trimmed <- region_iqr_sthlm[1:nrow(region_iqr_skane), ]

region_similarity_pre <- cbind(region_iqr_skane, region_iqr_sthlm_trimmed) |> 
  rename(sthlm_iqr = region_iqr_sthlm_trimmed)
```

```{r}
# Same for post AbsPQN2
region_iqr_skane_v2 <- median_iqr_skane_v2 |> 
  select(iqr_post) |> 
  rename(skane_iqr = iqr_post)

region_iqr_sthlm_v2 <- median_iqr_sthlm_v2 |> 
  select(iqr_post) |> 
  rename(sthlm_iqr = iqr_post)

# Different length Skane and Sthlm, trim some elements from the longer df
region_iqr_sthlm_trimmed_v2 <- region_iqr_sthlm_v2[1:nrow(region_iqr_skane_v2), ]

region_similarity_post <- cbind(region_iqr_skane_v2, region_iqr_sthlm_trimmed_v2) |> 
  rename(sthlm_iqr = region_iqr_sthlm_trimmed_v2)
```

Making plots
```{r}
# IQR comparison, pre
ggplot(region_similarity_pre, aes(x = skane_iqr, y = sthlm_iqr)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " IQR Skane", y = "IQR Stockholm",  title = "IQR similarity between regions - Samples", subtitle = "Pre AbsPQN2") +
  theme_minimal()

# IQR comparison, post
ggplot(region_similarity_post, aes(x = skane_iqr, y = sthlm_iqr)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = " IQR Skane", y = "IQR Stockholm",  title = "IQR similarity between regions - Samples", subtitle = "Post AbsPQN2") +
  theme_minimal()

```
IQR FOR PROTEINS
```{r}
median_iqr_prot <- function(df) {
  
  df <- df |> 
    select(-sample_id, -individual_id)
  
  medians <- apply(df, 2, median)
  
  # Compute IQR for each row
  iqrs <- apply(df, 2, IQR)
  
  # Combine medians and IQRs into a results dataframe
  results <- data.frame(median = medians, iqr = iqrs)

  return(results)
}

# Pre
median_iqr_skane <- median_iqr_prot(pre_skane) |> 
  rename(median_pre = median) |> 
  rename(iqr_pre = iqr) |> 
  rownames_to_column(var = "protein")
median_iqr_sthlm <- median_iqr_prot(pre_sthlm) |> 
  rename(median_pre = median) |> 
  rename(iqr_pre = iqr) |> 
  rownames_to_column(var = "protein")

# AbsPQN2
median_iqr_skane_v2 <- median_iqr_prot(post_skane_v2)|> 
  rename(median_post = median) |> 
  rename(iqr_post = iqr) |> 
  rownames_to_column(var = "protein")
median_iqr_sthlm_v2 <- median_iqr_prot(post_sthlm_v2) |> 
  rename(median_post = median) |> 
  rename(iqr_post = iqr) |> 
  rownames_to_column(var = "protein")

# Selecting IQR
region_iqr_skane <- median_iqr_skane |> 
  select(iqr_pre, protein) |> 
  rename(skane_iqr = iqr_pre)

region_iqr_sthlm <- median_iqr_sthlm |> 
  select(iqr_pre, protein) |> 
  rename(sthlm_iqr = iqr_pre)

# Different length Skane and Sthlm, trim some elements from the longer df
#region_iqr_sthlm_trimmed <- region_iqr_sthlm[1:nrow(region_iqr_skane), ]

region_similarity_pre <- left_join(region_iqr_skane, region_iqr_sthlm, by = "protein")


# Same for post AbsPQN2
region_iqr_skane_v2 <- median_iqr_skane_v2 |> 
  select(iqr_post, protein) |> 
  rename(skane_iqr = iqr_post)

region_iqr_sthlm_v2 <- median_iqr_sthlm_v2 |> 
  select(iqr_post, protein) |> 
  rename(sthlm_iqr = iqr_post)

region_similarity_post <- left_join(region_iqr_skane_v2, region_iqr_sthlm_v2, by = "protein")

# Calculate the correlation coefficient
cor_pre <- cor(region_similarity_pre$skane_iqr, region_similarity_pre$sthlm_iqr, use = "complete.obs")

# Plot with annotation
ggplot(region_similarity_pre, aes(x = skane_iqr, y = sthlm_iqr)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = "IQR Skane", y = "IQR Stockholm", title = "IQR similarity between regions - Proteins", subtitle = "Pre AbsPQN2") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_pre, 3)), hjust = 3, vjust = 2) +
  theme_minimal()


# Calculate the correlation coefficient
cor_post <- cor(region_similarity_post$skane_iqr, region_similarity_post$sthlm_iqr, use = "complete.obs")

# Plot with annotation
ggplot(region_similarity_post, aes(x = skane_iqr, y = sthlm_iqr)) +
  geom_point(color = "#4DBBD5FF", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#265073", linetype = "dashed") +
  labs(x = "IQR Skane", y = "IQR Stockholm", title = "IQR similarity between regions - Proteins", subtitle = "Post AbsPQN2") +
  annotate("text", x = Inf, y = Inf, label = paste("Correlation: ", round(cor_post, 3)), hjust = 3, vjust = 2) +
  theme_minimal()

```
