```{r}
library(tidyverse)
library(viridis)
library(FactoMineR)
library(factoextra)
library(OlinkAnalyze)
library(pheatmap)
library(umap)
library(ggforce)
```

## Loading data

```{r}
# Clinical data
clinical_skane <- read_csv(file.path("../clean_data/clinical_skane_filtered.csv"), show_col_types = FALSE)
clinical_sthlm <- read_csv(file.path("../clean_data/clinical_sthlm_filtered.csv"), show_col_types = FALSE)

# NPX data
qc_skane <- read_csv(file.path("../clean_data/npx_skane_filtered.csv"), show_col_types = FALSE) |>  # (qc = only qc)
  mutate(type = "qc")
v2_skane <- read_csv(file.path("../clean_data/norm_npx_skane_v2.csv"), show_col_types = FALSE) |>  # (v2 = abspqn2)
  mutate(type = "v2")
adj_skane <- read_csv(file.path("../clean_data/adjusted_npx_skane.csv"), show_col_types = FALSE) |>  # (adj = clin. adjusted)
  mutate(type = "adj")

qc_sthlm <- read_csv(file.path("../clean_data/npx_sthlm_filtered.csv"), show_col_types = FALSE) |> 
  mutate(type = "qc")
v2_sthlm <- read_csv(file.path("../clean_data/norm_npx_sthlm_v2.csv"), show_col_types = FALSE) |> 
  mutate(type = "v2")
adj_sthlm <- read_csv(file.path("../clean_data/adjusted_npx_sthlm.csv"), show_col_types = FALSE) |> 
  mutate(type = "adj")
```

With the "type" column, the different types of data can be plotted together, although they need to be put into the same data frame.

```{r}
comb_skane <- bind_rows(qc_skane, v2_skane, adj_skane)
comb_sthlm <- bind_rows(qc_sthlm, v2_sthlm, adj_sthlm)
```

# Density plots

Function for creating density plots for each of the proteins in the dataset.

```{r}
# Here we decide the number of plots per page
facets_per_page <- 25  # Adjust this number based on your needs

total_facets <- 159
total_pages <- ceiling(total_facets / facets_per_page)

# Function for histograms
plot_page <- function(page_number, facets_per_page, data) {
  
  long_data <- data |> 
    select(-sample_id) |> 
    pivot_longer(-type, values_to = "value", names_to = "name")

  start <- ((page_number - 1) * facets_per_page) + 1
  end <- min(start + facets_per_page - 1, total_facets)
  
  facet_names <- unique(long_data$name)[start:end]
  data_subset <- long_data |> filter(name %in% facet_names)
  
  plots <- ggplot(data_subset, aes(x = value, fill = type, group = type)) + 
    geom_histogram(alpha = 0.8) +
    facet_wrap(~ name, scales = "free") +
    theme_minimal() +
    scale_fill_npg() + 
    theme(legend.position = "none")
  
  return(plots)
}
```

### Skane density plots
```{r}
# Only qc data
pdf("../images/density_plots/qc_skane.pdf", paper = "a4")
# Loop through each page and plot the facets
walk(1:total_pages, ~ {
  print(plot_page(.x, facets_per_page, qc_skane))
})
# Close the PDF device
dev.off()

# AbsPQN2 data
pdf("../images/density_plots/v2_skane.pdf", paper = "a4")
# Loop through each page and plot the facets
walk(1:total_pages, ~ {
  print(plot_page(.x, facets_per_page, v2_skane))
})
# Close the PDF device
dev.off()

# Adjusted data
pdf("../images/density_plots/adj_skane.pdf", paper = "a4")
# Loop through each page and plot the facets
walk(1:total_pages, ~ {
  print(plot_page(.x, facets_per_page, adj_skane))
})
# Close the PDF device
dev.off()

# All in the same plot
pdf("../images/density_plots/comb_skane.pdf", paper = "a4")
# Loop through each page and plot the facets
walk(1:total_pages, ~ {
  print(plot_page(.x, facets_per_page, comb_skane))
})
# Close the PDF device
dev.off()
```
### Stockolm density plots
```{r}
# Only qc data
pdf("../images/density_plots/qc_sthlm.pdf", paper = "a4")
# Loop through each page and plot the facets
walk(1:total_pages, ~ {
  print(plot_page(.x, facets_per_page, qc_sthlm))
})
# Close the PDF device
dev.off()

# AbsPQN2 data
pdf("../images/density_plots/v2_sthlm.pdf", paper = "a4")
# Loop through each page and plot the facets
walk(1:total_pages, ~ {
  print(plot_page(.x, facets_per_page, v2_sthlm))
})
# Close the PDF device
dev.off()

# Adjusted data
pdf("../images/density_plots/adj_sthlm.pdf", paper = "a4")
# Loop through each page and plot the facets
walk(1:total_pages, ~ {
  print(plot_page(.x, facets_per_page, adj_sthlm))
})
# Close the PDF device
dev.off()

# All in the same plot
pdf("../images/density_plots/comb_sthlm.pdf", paper = "a4")
# Loop through each page and plot the facets
walk(1:total_pages, ~ {
  print(plot_page(.x, facets_per_page, comb_sthlm))
})
# Close the PDF device
dev.off()
```

# Correlation plots / Heatmaps
```{r}
# Skane
corr_qc_skane <- qc_skane |> 
  select(-sample_id, -type) |> 
  cor(method = "spearman", use = "complete.obs") |> 
  as_tibble(rownames = "protein") |> 
  column_to_rownames(var = "protein")

corr_v2_skane <- v2_skane |> 
  select(-sample_id, -type) |> 
  cor(method = "spearman", use = "complete.obs") |> 
  as_tibble(rownames = "protein") |> 
  column_to_rownames(var = "protein")

corr_adj_skane <- adj_skane |> 
  select(-sample_id, -type) |> 
  cor(method = "spearman", use = "complete.obs") |> 
  as_tibble(rownames = "protein") |> 
  column_to_rownames(var = "protein")

# Stockholm
corr_qc_sthlm <- qc_sthlm |> 
  select(-sample_id, -type) |> 
  cor(method = "spearman", use = "complete.obs") |> 
  as_tibble(rownames = "protein") |> 
  column_to_rownames(var = "protein")

corr_v2_sthlm <- v2_sthlm |> 
  select(-sample_id, -type) |> 
  cor(method = "spearman", use = "complete.obs") |> 
  as_tibble(rownames = "protein") |> 
  column_to_rownames(var = "protein")

corr_adj_sthlm <- adj_sthlm |> 
  select(-sample_id, -type) |> 
  cor(method = "spearman", use = "complete.obs") |> 
  as_tibble(rownames = "protein") |> 
  column_to_rownames(var = "protein")

```


With the correlation matrices created, we can plot them in correlation heatmaps.

```{r}
custom_col = colorRampPalette(c("blue", "white", "red"))(200) # Controls the color palette
breaks = seq(-1, 1, length.out = length(custom_col) + 1)



# Skane PDF
pdf("../images/heatmaps/skane_correlation_plots.pdf", width = 25, height = 25)

pheatmap(corr_qc_skane, 
         cluster_rows = TRUE,    # Cluster rows (variables)
         cluster_cols = TRUE,    # Cluster columns (variables)
         color = custom_col,  # Color palette
         breaks = breaks,
         main = "qc_skane - Clustered Correlation Heatmap")
pheatmap(corr_v2_skane, 
         cluster_rows = TRUE,    # Cluster rows (variables)
         cluster_cols = TRUE,    # Cluster columns (variables)
         color = custom_col,  # Color palette
         breaks = breaks,
         main = "v2_skane - Clustered Correlation Heatmap")
pheatmap(corr_adj_skane, 
         cluster_rows = TRUE,    # Cluster rows (variables)
         cluster_cols = TRUE,    # Cluster columns (variables)
         color = custom_col,  # Color palette
         breaks = breaks,
         main = "adj_skane - Clustered Correlation Heatmap")

# Close the PDF device
dev.off()

# Stockholm PDF
pdf("../images/heatmaps/sthlm_correlation_plots.pdf", width = 25, height = 25)

custom_col = colorRampPalette(c("blue", "white", "red"))(200) # Controls the color palette

pheatmap(corr_qc_sthlm, 
         cluster_rows = TRUE,    # Cluster rows (variables)
         cluster_cols = TRUE,    # Cluster columns (variables)
         color = custom_col,  # Color palette
         breaks = breaks,
         main = "qc_sthlm - Clustered Correlation Heatmap")
pheatmap(corr_v2_sthlm, 
         cluster_rows = TRUE,    # Cluster rows (variables)
         cluster_cols = TRUE,    # Cluster columns (variables)
         color = custom_col,  # Color palette
         breaks = breaks,
         main = "v2_sthlm - Clustered Correlation Heatmap")
pheatmap(corr_adj_sthlm, 
         cluster_rows = TRUE,    # Cluster rows (variables)
         cluster_cols = TRUE,    # Cluster columns (variables)
         color = custom_col,  # Color palette
         breaks = breaks,
         main = "adj_sthlm - Clustered Correlation Heatmap")

# Close the PDF device
dev.off()
```

```{r}
# Converting to long format for easier management
long_corr_adj_skane <- corr_adj_skane |> 
  as_tibble(rownames = "Protein1") |> 
  pivot_longer(-Protein1, names_to = "Protein2", values_to = "Correlation") |> 
  filter(!(Protein1 == Protein2)) |> # Removing diagonal (self correlation) 
  filter(Correlation >= 0.5)

long_corr_adj_sthlm <- corr_adj_sthlm |> 
  as_tibble(rownames = "Protein1") |> 
  pivot_longer(-Protein1, names_to = "Protein2", values_to = "Correlation") |> 
  filter(!(Protein1 == Protein2))
```

```{r}
skane_pheatmap <- pheatmap(corr_adj_skane, cluster_rows = TRUE, cluster_cols = TRUE)
dev.off() # Preventing plot output
clusters <- cutree(skane_pheatmap$tree_row, k = 4)

# Identifying the cluster which is highly correlated in the plot
protein_name <- "protein_name"  
protein_cluster <- clusters[protein_name]

# Saving the protein names found in this cluster
skane_clusters <- names(clusters)[clusters == 3] |> 
  as_tibble()
write.csv(skane_clusters, "skane_clustered_proteins")

sthlm_pheatmap <- pheatmap(corr_adj_sthlm, cluster_rows = TRUE, cluster_cols = TRUE)
dev.off() # Preventing plot output
clusters <- cutree(sthlm_pheatmap$tree_row, k = 4)

# Identifying the cluster which is highly correlated in the plot
protein_name <- "protein"  
protein_cluster <- clusters[protein_name]

# Saving the protein names found in this cluster
sthlm_cluster <- names(clusters)[clusters == 3] |> 
  as_tibble()
write.csv(sthlm_cluster, "sthlm_clustered_proteins")
```

# PCA

Skane PCA grid
```{r}
# Adding x_BC
skane_x <- clinical_skane |> 
  select(x_BC, sample_id) |> 
  mutate(x_BC = as_factor(x_BC))

# Prepping for PCA using recipes
adj_skane_prep <- 
  recipe(~ ., data = adj_skane |> select(-type)) |>
  update_role(sample_id, new_role = "ID") |> # Ensures that sample_id isn't used in any calculations
  step_center(all_predictors()) |>  
  step_scale(all_predictors()) |> 
  step_pca(all_predictors(), num_comp = 3) |>  
  prep() |> 
  juice() |> 
  left_join(skane_x, by = "sample_id") |> 
  select(-sample_id)

ggplot(adj_skane_prep, aes(x = .panel_x, y = .panel_y, color = x_BC, fill = x_BC)) + 
  geom_point(alpha = 0.3, shape = 16, size = 1) + 
  geom_autodensity(alpha = 0.3) +
  facet_matrix(vars(-x_BC), layer.diag = 2, grid.y.diag = FALSE) +
  scale_color_npg() + 
  scale_fill_npg()
```

Stockholm PCA grid
```{r}
# Adding x_BC
sthlm_x <- clinical_sthlm |> 
  select(x_BC, sample_id) |> 
  mutate(x_BC = as_factor(x_BC))

# Prepping for PCA using recipes
adj_sthlm_prep <- 
  recipe(~ ., data = adj_sthlm |> select(-type)) |>
  update_role(sample_id, new_role = "ID") |>
  step_center(all_predictors()) |>  
  step_scale(all_predictors()) |> 
  step_pca(all_predictors(), num_comp = 3) |>  
  prep() |> 
  juice() |> 
  left_join(sthlm_x, by = "sample_id") |> 
  select(-sample_id)

ggplot(adj_sthlm_prep, aes(x = .panel_x, y = .panel_y, color = x_BC, fill = x_BC)) + 
  geom_point(alpha = 0.2, shape = 16, size = 0.5) + 
  geom_autodensity(alpha = 0.3) +
  facet_matrix(vars(-x_BC), layer.diag = 2, grid.y.diag = FALSE) +
  scale_color_npg() + 
  scale_fill_npg()
```

## UMAP

```{r}
# Run the UMAP
umap_skane <- adj_skane |> 
  select(-sample_id, -type) |> 
  umap()
umap_sthlm <- adj_sthlm |> 
  select(-sample_id, -type) |> 
  umap

umap_adj_skane <- adj_skane |> 
  left_join(skane_x, by = "sample_id")

umap_adj_sthlm <- adj_sthlm |> 
  left_join(sthlm_x, by = "sample_id")

# Add the UMAP variables to the NPX data frames
umap_adj_skane$UMAP1 <- umap_skane$layout[, 1]
umap_adj_skane$UMAP2 <- umap_skane$layout[, 2]

umap_adj_sthlm$UMAP1 <- umap_sthlm$layout[, 1]
umap_adj_sthlm$UMAP2 <- umap_sthlm$layout[, 2]
```

```{r}
# Clustering using kmeans
set.seed(123) # For reproducibility
clusters_skane <- kmeans(umap_skane$layout, centers = 3) 
clusters_sthlm <- kmeans(umap_sthlm$layout, centers = 3)

# Add cluster assignments to the data frames
umap_adj_skane$Cluster <- as.factor(clusters_skane$cluster) 
umap_adj_sthlm$Cluster <- as.factor(clusters_sthlm$cluster)

umap_adj_skane <- umap_adj_skane |> 
  left_join(clinical_skane, by = "sample_id")

umap_adj_sthlm <- umap_adj_sthlm |> 
  left_join(clinical_sthlm, by = "sample_id")
```

```{r}
ggplot(umap_adj_skane, aes(x = UMAP1, y = UMAP2, color = clinical_variable)) +
  geom_point(alpha = 0.8) + 
  theme_minimal() +
  labs(title = "UMAP Projection - Skane", x = "UMAP 1", y = "UMAP 2")

ggplot(umap_adj_sthlm, aes(x = UMAP1, y = UMAP2, color = clinical_variable)) +
  geom_point(alpha = 0.8) + 
  theme_minimal() +
  labs(title = "UMAP Projection - Stockholm", x = "UMAP 1", y = "UMAP 2")
```


